---
title: "Final Project"
editor: visual
---

## Group 1:

-   Hiba Awan

-   Nathania Stephens

# [Abstract]{.underline}

# [Introduction & Background]{.underline}

## Motivation/ Purpose

In 2023, there were over 30,000 arrests and close to 65,000 citations in Fairfax County. The Fairfax County boundaries, include areas such as Centreville, Chantilly, Herndon, Reston, Tysons Corner, McLean, Merrifield, George Mason, Annadale, Burke, Springfield, Alexandria, Lorton to name a few. If you live, work, or study in these areas then this project should be of interest to you. This project aims to inform Fairfax County patrons of crime information and hopefully provide some statistical insights that could be applicable.

## Goals/ Objectives

In order to provide relevant and insightful crime information, several different visualization methods were applied to help easily interpret and compare data. Statistical learning techniques were utilized to help understand statistical significantly factors and associations between variables. Since the data utilized for this project is largely categorical the project focuses on techniques such as Chi-Squared Test, *Logistic Regression, Decision Trees and Random Forest*.

# [Data]{.underline}

## Overview

## About the Data

Three datasets were pulled from the Fairfax County Police Department website. They covered arrest, citations, and warnings in the year 2023. For simplicity general definition are provided:

**Arrest** - When a person is taken into custody to answer for an offense or when there is a deprivation or restraint of a person's liberty in any significant way.

**Citation** - Formal notice issued by law enforcement officer for a violation of law, typically related to traffic laws or other minor offenses. Typically requiring a violator to appear in court or pay a fine.

**Warning** - When a violation, typically minor, has been made but an officer issues a warning rather than a citation.

The following attributes were key to the research conducted:

| Column Name | Data Type | Description                                       |
|-------------|-----------|---------------------------------------------------|
| Date        | Date      | Date of Violation                                 |
| Time        | Chr       | Time of Violation                                 |
| Offense     | Chr       | Description of Violation                          |
| Gender      | Chr       | Gender of Violator                                |
| Ethnicity   | Chr       | Hispanic or Non-Hispanic                          |
| District    | Chr       | Administrative area                               |
| Latitude    | Dbl       | Coordinates measuring north/ south of equator     |
| Longitude   | Dbl       | Coordinates measuring east/ west of prime meridian|
| Outcome     | Chr       | Result of violation, arrest, citation, or warning |


## Limitations and Assumptions
Due to the nature of the data available on the Fairfax County Police Department website, analysis was limited to qualitative techniques. The approach taken for the project focused on predicting through qualitative responses or classification. This means that each record pulled from the Fairfax County Police Department (FCPD) would be assigned to a category or class. 

While understanding local crime is the goal of this project, the data acquired only accounts for crime that was recorded by FCPD. It does not take into account crimes that were not report or any other crime that may have been reported through FCPD.

## Cleaning and Transformation

# [Research Questions]{.underline}

1)  Is there an association between gender and warnings?

2)  Are there other factors that determine if someone gets out of a "ticket"? OR Are you more likely to get a ticket at the end of the month (some believe that police officers have a monthly quota)

3)  

## Exploratory Analysis

To address each of these question, first exploratory analysis should be done to gain an understanding and summary of the crime metrics for Fairfax County. This includes understanding what type of crimes occurred the most and where.

General crime Mapping the arrest data for a geospatial visual of where arrest occur.l

```{r}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(leaflet)
library(usmap)
library(sf)
library(readr)
library(scales)
library(tidyr)

Arrest2023 <- read_csv("2023_arrest_data.csv", 
                       col_types = cols(ArresteeNumber = col_skip(), 
                                        ArrestDate = col_date(format = "%m/%d/%Y"), 
                                        ArrestTime = col_character(), WEB_ADDRESS = col_skip(), 
                                        PHONE_NUMBER = col_skip(), NAME = col_skip()))

## Remove any outliers by setting min/max lat and long.
min_lat = 38.6
max_lat = 39.0
min_lon = -77.6
max_lon = -77.0

## Using all data but excluding any coordinates that are not within coordinates
arrest_data_clean = Arrest2023 %>%
  filter(
    Latitude >= min_lat,
    Latitude <= max_lat,
    Longitude >= min_lon,
    Longitude <= max_lon
  )

## Include district boundaries by reading in shape file downloaded from Fairfax County site
district_boundaries = st_read("Supervisor_Districts/Supervisor_Districts.shp")

## Transform to WGS84 projection to match basemap within leaflet
if (st_crs(district_boundaries)$epsg != 4326) {
  district_boundaries = st_transform(district_boundaries, 4326)
}
```

```{r}
#| echo: false
#| message: false
#| warning: false
## Using leaflet to map the Arrest data and include district layers
leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(
    data = district_boundaries,
    fillOpacity = 0.5,
    color = "black",
    weight = 1,
    popup = ~paste("District:", DISTRICT)
  ) %>%
  addCircleMarkers(
    data = arrest_data_clean,
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    color = "darkred",
    stroke = FALSE,
    fillOpacity = 0.4,
    popup = ~paste(
      "Case Number: ", CaseNumber, "<br>",
      "Arrest Type: ", IBRDescription)
  )
```

```{r}

```

Next we look at the **Top 10 Arrest Type** by Incident Based Reporting (IBR) codes.

```{r}
#| echo: false
#| warning: false
# By IBRCode
arrest_count = Arrest2023 %>%
  group_by(IBRCode) %>%
  summarise(Count = n()) %>%
  ungroup() %>%
  arrange(desc(Count))

# Get top 10
top_10_arrest_IBR = head(arrest_count, 10)

# Add codes - General Description
IBRCodeDetails = read.csv("IBRcodes.csv")

# Reassign the codes
top_10_arrest_decoded = top_10_arrest_IBR %>%
  left_join(IBRCodeDetails, by = c("IBRCode" = "CODE")) %>%
  select(IBRCode, OFFENSE, everything())

top_10_arrest_cleaned = top_10_arrest_decoded %>%
  distinct(IBRCode, OFFENSE, .keep_all = TRUE)

# Decode for cleaner results
arrest_chart = ggplot(top_10_arrest_cleaned, aes(x = reorder(OFFENSE, Count),
                                              y = Count)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = "Top 10 Arrest Type by IBR decoded",
       x = "Offense by IBR Code",
       y = "Number of Arrests") +
  theme_grey() +
  geom_text(aes(label = Count), hjust = -0.1, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

arrest_chart

```

Next examining the **Top 10 Citations**

**Warning Versus Citation** Next an examination of warning versus citation will be observed... This will help understand what different factors could play into getting a warning or a citation.

```{r}
#| echo: false
#| warning: false
library(readr)
library(lubridate)

warnings = read_csv("2023_warning_data.csv", 
                         col_types = cols(Warnings_Date = col_date(format = "%m/%d/%Y"), 
                                          WEB_ADDRESS = col_skip(), PHONE_NUMBER = col_skip(), 
                                          NAME = col_skip()))

citations = read_csv("2023_citation_data.csv", 
                                col_types = cols(Date = col_date(format = "%m/%d/%Y"), 
                                                 WEB_ADDRESS = col_skip(), PHONE_NUMBER = col_skip(), 
                                                 NAME = col_skip()))

# Rename some columns
citations = citations %>%
  rename(ViolationDate = Date)

# change Gender to sex in warnings and change date column name
warnings = warnings %>%
  rename(Sex = Gender)

warnings = warnings %>%
  rename(ViolationDate = Warnings_Date)

# Adjust Citations and prepare for Merge
# Assumption that ID is the officer's ID
citations_processed = citations %>%
  mutate(
    outcome = "Citation",
    Gender = case_when(
      Sex == "M" ~ "Male",
      Sex == "F" ~ "Female",
      TRUE ~ "Other/Unknown"
    ),
    Year = year(ViolationDate),
    Month = month(ViolationDate),
    DayOfMonth = day(ViolationDate),
    Time = parse_date_time(Time, "HM"),
    data_type = "Citation"
  ) %>%
  select(
    outcome, Gender, Year, Month, DayOfMonth, Time, Offense_Description = Charge,
    District = DISTRICT, Race, Ethnicity, Latitude, Longitude, OfficerID = ID, data_type
  )

# Adjust Warnings and prepare for Merge
warnings_processed = warnings %>%
  mutate(
    outcome = "Warning",
    Gender = case_when(
      Sex == "M" ~ "Male",
      Sex == "F" ~ "Female",
      TRUE ~ "Other/Unknown"
    ),
    Year = year(ViolationDate),
    Month = month(ViolationDate),
    DayOfMonth = day(ViolationDate),
    Time = parse_date_time(Time, "HM"),
    data_type = "Warning"
  ) %>%
  select(
    outcome, Gender, Year, Month, DayOfMonth, Time, Offense_Description, District = DISTRICT, Race, 
    Ethnicity, Latitude = Lat, Longitude = Long, OfficerID = Officer_ID, data_type
  )

# Combined for ultimate Data coordination!
combined_wc = bind_rows(citations_processed, warnings_processed)

# Add ultimate binary outcome! 0 = Citation, 1 = Warning/ Got out of ticket
combined_wc = combined_wc %>%
  mutate(
    BinaryOutcome = ifelse(outcome == "Warning", 1,0)
  )

## Change to Title Case for District Names
combined_wc$District = tools::toTitleCase(tolower(combined_wc$District))

## Examining Unverified data
## After examination, unverified only makes up 0.0143 or 1.43% of the data set, so we will remove
## because it is a very small portion of the total proportion. 
combined_wc %>%
  count(District) %>%
  mutate(Proportion = n / sum(n)) %>%
  arrange(desc(n))

## Filter out Unverified and NA
combined_wc = combined_wc %>%
  filter(District != "Unverified")

combined_wc = combined_wc %>%
  filter(!is.na(District))

## Filter out Other/Unknown Gender
combined_wc_mf = combined_wc %>%
  filter(Gender != "Other/Unknown")

## Now for some visuals: Gender Chart
## Examining the proportion of stops resulting in a Warning Vs Citation
## the Warning rate is the proportion of incidents that are warnings.
gender_warning_rate = combined_wc_mf %>%
  group_by(Gender) %>%
  summarise(
    Total_Incidents = n(),
    Warning_Rate = mean(BinaryOutcome)
  ) %>%
  ungroup()

gender_chart = ggplot(gender_warning_rate,
                      aes(x = Gender, y = Warning_Rate, fill = Gender)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::percent(Warning_Rate, accuracy = 0.1)),
            vjust = -0.5, size = 5) +
  scale_y_continuous(labels = scales::percent, limits = c(0, max(gender_warning_rate$Warning_Rate) * 1.1)) +
  labs(
    title = "Warning Rate by Gender",
    subtitle = "Proportion of stops resulting in a Warning (vs Citation)",
    x = "Gender",
    y = "Warning Rate"
  ) + theme_gray() + theme(plot.title = element_text(hjust = 0.5)) + theme(plot.subtitle = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "skyblue"))

gender_chart
```

```{r}
#| echo: false
#| warning: false
## Now the Chi-Squared Test starting with the Contingency Table
contingency_tbl = combined_wc_mf %>%
  filter(Gender %in% c("Male", "Female")) %>%
  select(Gender, BinaryOutcome) %>%
  table()

contingency_tbl

chi_sq_results = chisq.test(contingency_tbl)

chi_sq_results
```

# [Conclusion]{.underline}

# [References]{.underline}
