---
title: "Helpful R Info"
editor: visual
---

# Data types

## Identify data types

Is it a logical (Boolean), string, etc?

```{r}
x = 5
class(x)
```

Or use is....?

```{r}
is.numeric(x)
```

## Change a data type

```{r}
x = as.character(x)
x
```

```{r}
is.numeric(x)
```

```{r}
class(x)
```

# TRUE or FALSE?
Binary Comparison, is it true or false?

Using variable **a** and **b**:

```{r}
a = 5
b = 11
```

-   Is **a** equal to **b**?

```{r}
a == b
```

-   Is **a** less than **b**?

```{r}
a < b
```

-   Is **a** greater than **b**?

```{r}
a > b
```

-   Is **a** less than or equal to **b**?

```{r}
a <= b
```

-   Is **a** greater than or equal to **b**?

```{r}
a >= b
```

-   Is **a** not equal to **b**?

```{r}
a != b
```

# Basic Operations

Floor will round down always.

```{r}
x = 3.2
floor(x)
```

Ceiling will round up always.

```{r}
ceiling(x)
```

# Math?!

Square root √

```{r}
sqrt(x)
```

Exponent
```{r}
3^2
```

# Variables

Assign multiple variables at once.

```{r}
a = b = 11
a
b
```

```{r}
b
```

## Remove a Variable

```{r}
rm(b)
```

#Vectors Collection of elements, like a list.

```{r}
myvector = c('apple', 'banana', 'orange', 'pineapple')
myvector
```

# What does the colon ":" do again?

Create consecutive numbers without writing them all out using the ":".

```{r}
z = c(1:10)
z
```

Decreasing order
```{r}
10:2
```
# Data Frames
Creating data frames
w = 2*z
```{r}
w = 2*z
y = c("a","b","c","d","e","f","g","h","i","j")
```
# Directory
## Get current directory
Get the current working directory you are in.
```{r, eval=FALSE}
getwd()
```
## Change directory
Change your directory another location
```{r, eval=FALSE}
setwd("EnTeR NeW PaTh") # use // or \ for windows OS
```

# Explore the data
## View
To see a data frame use "View"
```{r, eval=FALSE}
View(penguins)
```
## Summary
Summary will provide a summary of the rows, including NA rows.
```{r}
summary(penguins)
```
## Glimpse
Use "glimpse" to see the number of rows, columns, and each of the column names with their data type.
```{r, eval=FALSE}
glimpse(penguins)
```

# Creating a ggplot
Make sure to run the library for ggplot
```{r}
library(tidyverse)
ggplot(penguins, aes(x = flipper_len, y = body_mass)) +
  geom_point()
```

