---
title: "Helpful R Info"
editor: visual
---

# Data types

## Identify data types

Is it a logical (Boolean), string, etc?

```{r}
x = 5
class(x)
```

Or use is....?

```{r}
is.numeric(x)
```

## Change a data type

```{r}
x = as.character(x)
x
```

```{r}
is.numeric(x)
```

```{r}
class(x)
```

# Binary Comparison: Is it TRUE or FALSE?

Using variable **a** and **b**:

```{r}
a = 5
b = 11
```

-   Is **a** equal to **b**?

```{r}
a == b
```

-   Is **a** less than **b**?

```{r}
a < b
```

-   Is **a** greater than **b**?

```{r}
a > b
```

-   Is **a** less than or equal to **b**?

```{r}
a <= b
```

-   Is **a** greater than or equal to **b**?

```{r}
a >= b
```

-   Is **a** not equal to **b**?

```{r}
a != b
```

# Basic Operations

Floor will round down always.

```{r}
x = 3.2
floor(x)
```

Ceiling will round up always.

```{r}
ceiling(x)
```

# Variables

Assign multiple variables at once.

```{r}
a = b = 11
a
b
```

```{r}
b
```

## Remove a Variable

```{r}
rm(b)
```

#Vectors
Collection of elements, like a list.
```{r}
myvector = c('apple', 'banana', 'orange', 'pineapple')
myvector
```

# What does the colon ":" do again?
Create consecutive numbers without writing them all out using the ":".
```{r}
z = c(1:10)
z
```




